HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model

### 2020-02-19 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add DREAM across all cohorts

---
title: "HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
author: "Kelsey Montgomery and Thanneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                          parentId ="syn21435176",
                                          folderName = 'HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
```

@@

library(synapser)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

@@

# Parse gene lengths
gene_lengths = COUNT[,c("Geneid", "Length")]
COUNT$Length <- NULL
gene_lengths <- rename(gene_lengths, gene_id = Geneid)

@@

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>% 
  dplyr::rename(Tissue = `rnaSeq_dissection:Brain_Region`,
                PMI = `PMI_(in_hours)`,
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                GenotypeExclude = `SNP_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                IsolationBatch = `rnaSeq_isolation:RNA_Isolation_Batch`, #isolation batch to replace ribozero batch
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Tissue, Reported_Gender, Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, IsolationBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "MDD", Other = "BP", Control = "Control", SCZ = "SCZ"))

@@

METADATA %>% 
  group_by(Dx) %>% 
  summarise(count = n()) %>% 
  kable()

@@

```{r Reported_Gender_plots}
COUNT$gene_id = rownames(COUNT) 
REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
  left_join(COUNT) %>%
  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
  mutate(value = log(value)) %>%
  rename(`counts(log)`= value) %>% 
  rename(SampleID = item) %>%
  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
  rename(`Reported Gender` = Reported_Gender)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, "SampleID")) %>%
    tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID) %>% 
  dplyr::mutate(label = ifelse((label %in% outliers), label, NA))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Tissue, size=ageOfDeath))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% outliers),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
p

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA HBCC ACC-STG Cohort") + 
  guides(colour = guide_legend(nrow = 2),
         shape = guide_legend(nrow = 2),
         size = guide_legend(nrow = 2))
p
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)

@@

p <- ggplot(plotdata, aes(MDS1,MDS2))
p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS HBCC ACC-STG Cohort") + 
  guides(colour = guide_legend(nrow = 2),
         shape = guide_legend(nrow = 2),
         size = guide_legend(nrow = 2))
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Reported_Gender", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
tmp = NEW.COUNTS
tmp[is.na(tmp)] = 0
VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)

# Estimate consensus correlation coefficient
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(tmp, 
                            design=DM1$design, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Tissue, size=ageOfDeath))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```

@@

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueOther_STG-Dx.TissueControl_STG",
                                     "Dx.TissueSCZ_STG-Dx.TissueControl_STG",
                                     "Dx.TissueSCZ_STG-Dx.TissueOther_STG"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:3, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('gene_id') %>%
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) = colnames(contrast)

@@

# Code
CODE <- Folder(name = "HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)

@@

# Store enrichment results
write.table(enrichment.results, file = 'CMC_HBCC_ACC-STG_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_HBCC_ACC-STG_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'enrichResults'
synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

>>>>

---
title: "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
author: "Kelsey Montgomery and Thanneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                          parentId ="syn21435176",
                                          folderName = 'HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
```

@@

library(variancePartition)
library(synapser)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

@@

# Parse gene lengths
gene_lengths = COUNT[,c("Geneid", "Length")]
COUNT$Length <- NULL
gene_lengths <- dplyr::rename(gene_lengths, gene_id = Geneid)

@@

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>% 
  dplyr::rename(Tissue = `rnaSeq_dissection:Brain_Region`,
                PMI = `PMI_(in_hours)`,
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                GenotypeExclude = `SNP_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                IsolationBatch = `rnaSeq_isolation:RNA_Isolation_Batch`, #isolation batch to replace ribozero batch
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, IsolationBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "MDD", Other = "BP", Control = "Control", SCZ = "SCZ"))

@@

METADATA %>% 
  group_by(Dx) %>% 
  dplyr::summarise(count = n()) %>% 
  kable()

@@

```{r Reported_Gender_plots}
COUNT$gene_id = rownames(COUNT) 
REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
  left_join(COUNT) %>%
  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
  dplyr::mutate(value = log(value)) %>%
  dplyr::rename(`counts(log)`= value) %>% 
  dplyr::rename(SampleID = item) %>%
  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
  dplyr::rename(`Reported Gender` = Reported_Gender)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, "SampleID")) %>%
  dplyr::mutate(label = SampleID) %>% 
  dplyr::mutate(label = ifelse((label %in% outliers), label, NA))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
p <- p + theme_bw() %+replace% theme(legend.position="right")
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p
# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID'))
# p = ggplot(tmp1 %>%
#              dplyr::filter(SampleID %in% indToRemove),
#            aes(x = logCPM, color = SampleID)) + geom_density() 
# p = p + theme_bw() %+replace% theme(legend.position = 'none') + facet_grid(Dx~., scale = 'free')
# p

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA MSSM/Penn/Pitt Cohort") + 
  guides(colour = guide_legend(nrow = 2),
         shape = guide_legend(nrow = 2),
         size = guide_legend(nrow = 2))
p
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)

@@

p <- ggplot(plotdata, aes(MDS1,MDS2))
p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS MSSM/Penn/Pitt Cohort") + 
  guides(colour = guide_legend(nrow = 2),
         shape = guide_legend(nrow = 2),
         size = guide_legend(nrow = 2))
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Reported_Gender", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

# Post adjusted formula
random_effect <- "Individual_ID"
formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")

# Estimate voom weights with DREAM
tmp = NEW.COUNTS
tmp[is.na(tmp)] = 0
geneExpr = DGEList(tmp)
geneExpr = edgeR::calcNormFactors(geneExpr)

VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
                                                               formula = formula,
                                                               data = COVARIATES)

# CQN NA counts to 0 counts for input to dream()
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
VOOM.GENE_EXPRESSION$E = tmp
ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
                                        formula = formula,
                                        data = COVARIATES,
                                        computeResiduals = TRUE)
# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx.Tissue, scales = 'free_y')
p
```

@@

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")
L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueOther_ACC", "Dx.TissueControl_ACC"))
L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueControl_ACC"))
L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueOther_ACC"))
L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueOther_STG", "Dx.TissueControl_STG"))
L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueControl_STG"))
L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueOther_STG"))
L = cbind(L1, L2, L3, L4, L5, L6)
contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
              glue::glue_collapse(names(L2[L2 != 0]), "-"),
              glue::glue_collapse(names(L3[L3 != 0]), "-"),
              glue::glue_collapse(names(L4[L4 != 0]), "-"),
              glue::glue_collapse(names(L5[L5 != 0]), "-"),
              glue::glue_collapse(names(L6[L6 != 0]), "-"))
# Visualize contrast matrix
plotContrasts(L)

# Fit contrast
FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
                                     formula = formula,
                                     data = COVARIATES,
                                     L = L)
# Get differential expression
DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>%
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) <- contrast

@@

# Code
CODE <- Folder(name = "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)

@@

# Store enrichment results
write.table(enrichment.results, file = 'CMC_HBCC_ACC-STG_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_HBCC_ACC-STG_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'enrichResults'
synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
```

|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

### 2020-02-19 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add DREAM across all cohorts

knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                          parentId ="syn21435176",
                                          folderName = 'HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
```

@@

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
tmp = NEW.COUNTS
tmp[is.na(tmp)] = 0
VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)

# Estimate consensus correlation coefficient
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(tmp, 
                            design=DM1$design, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

@@

### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("DxOther-DxControl",
                                     "DxSCZ-DxControl",
                                     "DxSCZ-DxOther"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:3, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('gene_id') %>%
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths %>%
              rownameToFirstColumn('gene_id'))
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

# Code
CODE <- Folder(name = "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)

@@

# Store enrichment results
write.table(enrichment.results, file = 'CMC_HBCC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_HBCC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'enrichResults'
synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

>>>>

knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                          parentId ="syn21435176",
                                          folderName = 'HBCC - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
```

@@

# Post adjusted formula
random_effect <- "Individual_ID"
formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")

# Estimate voom weights with DREAM
tmp = NEW.COUNTS
tmp[is.na(tmp)] = 0
geneExpr = DGEList(tmp)
geneExpr = edgeR::calcNormFactors(geneExpr)

VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
                                                               formula = formula,
                                                               data = COVARIATES)

# Fit linear model using new weights and new design
# CQN NA counts to 0 counts for input to dream()
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
VOOM.GENE_EXPRESSION$E = tmp
ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
                                        formula = formula,
                                        data = COVARIATES,
                                        computeResiduals = TRUE)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

@@

### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
 varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
   ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
```

@@

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")

L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueOther_DLPFC", "Dx.TissueControl_DLPFC"))
L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueControl_DLPFC"))
L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueOther_DLPFC"))
L = cbind(L1, L2, L3)
contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
              glue::glue_collapse(names(L2[L2 != 0]), "-"),
              glue::glue_collapse(names(L3[L3 != 0]), "-"))

# Visualize contrast matrix
plotContrasts(L)

#Fit contrast
FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
                                     formula = formula,
                                     data = COVARIATES,
                                     L = L)
# Get differential expression
DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>%
    left_join(backgroundGenes)
}, FIT.CONTR) 
 names(DE) <- contrast
DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths %>%
              rownameToFirstColumn('gene_id'))
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

# Code
CODE <- Folder(name = "HBCC - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)

@@

# Store enrichment results
write.table(enrichment.results, file = 'CMC_HBCC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_HBCC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'enrichResults'
synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
```

|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |
